---
title: "Lesson 6: Do Diets Really Work? (continued)"
author: "by Anny-Claude Joseph"
date: "Last compiled on `r format(Sys.time(), '%d %B, %Y')`"
output:
  html_document:
    df_print: paged
---

```{r message=FALSE, echo=TRUE, include=FALSE, warning=FALSE}
knitr::opts_chunk$set(echo = TRUE)
colorize <- function(x, color) {
  if (knitr::is_latex_output()) {
    sprintf("\\textcolor{%s}{%s}", color, x)
  } else if (knitr::is_html_output()) {
    sprintf("<span style='color: %s;'>%s</span>", color, 
      x)
  } else x
}

```
<br>

#### Background
Recall that in Lesson 5 we investigated the impact diet (Vegetarian, Keto, and None) has on weight loss (kg lost). 

```{r Setup, warning=FALSE, message=FALSE}
library(tidyverse)

## Load data
diet.dat <-read_csv('C:/Users/annyclaude.joseph/OneDrive - West Point/Joseph_Teaching/AY22-1/MA376/Lessons/WeightLoss.csv')
#glimpse(diet.dat)
```

```{r F-test}
# One way ANOVA
diet.aov <- aov(wloss~diet, data = diet.dat)
summary(diet.aov)
```

Based on the results of the F-test we concluded that the observed differences in mean weight loss among the diets is more than we would expect by chance alone. However, knowing there is a statistically significant association between diet and average kilograms lost does not tell us the entire story. 


(@) What else might we want to know? `r colorize("*Which population means are different? How much do they differ by?*", "blue")`

To figure this out, we conduct *post-hoc analyses*. Note! We only do post-hoc analyses when the F-test is statistically significant to protect against an inflated experiment-wise Type I error rate.
**Ask them what Type I error is? Note that experiment-wise Type I error rate increases as the number of comparison increases.**

<br>

#### A. All Pairwise Comparisons
For this method we compare average kilograms lost for all possible diet pairs. There are different methods for this type of pairwise comparison, but our textbook uses the formula:
$$\small \text{difference in means } (\bar{y}_i - \bar{y}_j) \pm \text{(Multiplier) (SE of Residuals)}  \times \sqrt{\tfrac{1}{n_i}+\tfrac{1}{n_j}}$$

(@) For 95% confidence intervals the multiplier is approximately 2. Why? `r colorize("The Empirical Rule (Page 89)! For a symmetric mound-shaped distribution approximately 68% of the values lie within one standard deviation of the mean; approximately 95% lie within two standard deviations of the mean", "blue")`. 

<br>
There are other methods to compare the mean for one population to the mean of another, e.g., Bonferroni corrections and Tukey's Honest Significant Difference (TukeyHSD).

**Remind them that when a confidence interval includes 0 (null hypothesized value), we do not have enough evidence to conclude there is a difference in the mean weight loss for the two diets.**

(@) Calculate the 95% confidence intervals for each pair using TukeyHSDs. For which groups can we conclude there is a statistically significant difference in the average number of kilograms lost? `r colorize("Vegetarian-Keto, Vegetarian-None", "blue")`.

```{r Calculate all pairwise comparisons}
TukeyHSD(diet.aov)
```
<br>

(@) Interpret one of the significant confidence intervals.

<br>

#### B. t- Confidence Intervals for Each Population Mean
We can also calculate a confidence interval for each treatment group (diet group) using the formula:

$$\small \bar{y}_i \pm \text{(Multiplier)} \times \frac{\text{SE of Residuals}}{\sqrt{n_i}}$$

(@) Which groups have statistically significant population mean weight loss?  `r colorize("*Since zero is not in the confidence intervals for any of the diet groups, average weight loss is statistically significant for every diet.*", "blue")`

```{r 95% CI For each mean}
# Fit the Multiple Means Model
multi.lm <- lm(wloss ~ 0 + diet, data = diet.dat)
confint(multi.lm)                    
```

(@) Interpret one of the significant confidence intervals.

<br>

#### C. Prediction Intervals for a New Observational Unit
A prediction interval is a range of values that predicts the value of a new observation based on the existing model. We can calculate a prediction interval using the formula:
$$\small \bar{y_i }\pm \text{(Multiplier) (SE of Residuals)} \sqrt{1 + \tfrac{1}{n_i}}$$
**Note the validity conditions for prediction intervals on the bottom of page 96**

(@) Estimate a prediction interval for average kilograms lost for an individual who is on the Keto diet. Interpret the prediction interval. `r colorize("*We are 95% certain that a (future/new) individual who is on the Keto diet will lose between -3.9 and 12.5 kilos.*", "blue")`

```{r, Prediction interval}
newobs.data <- data.frame(diet='Keto')
predict(multi.lm, newobs.data, interval = 'prediction')  
```

(@) When we are making a prediction where is our 'uncertainty' coming from?`r colorize("*Prediction intervals take into account individual variations in weight loss and variations in weight loss between diets.*", "blue")` 

**This is why prediction intervals tend to be wider that confidence intervals which only take into account the sample to sample variation in the sample mean.**
<br>
<br>
