---
title: "Lesson 6: Do Diets Really Work? (continued)"
output:
  html_document:
    df_print: paged
editor_options:
  chunk_output_type: console
---
```{r, include=FALSE}
knitr::opts_chunk$set(
	echo = TRUE,
	message = FALSE,
	warning = FALSE)
```
<br>

#### Background
Recall that in Lesson 5 we investigated the impact diet (Vegetarian, Keto, and None) has on weight loss (kg lost). 

```{r Setup, warning=FALSE, message=FALSE}
library(tidyverse)

## Load data
diet.dat <-read_csv('/WeightLoss.csv')
#glimpse(diet.dat)
```

```{r F-test}
# One way ANOVA
diet.aov <-  
```

Based on the results of the F-test we concluded that the observed differences in mean weight loss among the diets is more than we would expect by chance alone. However, knowing there is a statistically significant association between diet and average kilograms lost does not tell us the entire story. 

(@) What else might we want to know?  

<br>
To figure this out, we conduct *post-hoc analyses*. Note! We only do post-hoc analyses when the F-test is statistically significant to protect against an inflated experiment-wise Type I error rate.
 
<br>

#### A. All Pairwise Comparisons
For this method we compare average kilograms lost for all possible diet pairs. There are different methods for this type of pairwise comparison, but our textbook uses the formula:
$$\small \text{difference in means } (\bar{y}_i - \bar{y}_j) \pm \text{(Multiplier) (SE of Residuals)}  \times \sqrt{\tfrac{1}{n_i}+\tfrac{1}{n_j}}$$

(@) For 95% confidence intervals the multiplier is approximately 2. Why? 

<br>
There are other methods to compare the mean for one population to the mean of another, e.g., Bonferroni corrections and Tukey's Honest Significant Difference (TukeyHSD).

<br>

(@) Calculate the 95% confidence intervals for each pair using TukeyHSDs. For which groups can we conclude there is a statistically significant difference in the average number of kilograms lost?  

```{r Calculate all pairwise comparisons}
 
```
<br>

(@) Interpret one of the significant confidence intervals.

<br>

#### B. t- Confidence Intervals for Each Population Mean
We can also calculate a confidence interval for each treatment group (diet group) using the formula:

$$\small \bar{y}_i \pm \text{(Multiplier)} \times \frac{\text{SE of Residuals}}{\sqrt{n_i}}$$

(@) Which groups have statistically significant population mean weight loss? 

```{r 95% CI For each mean}
# First fit the Multiple Means Model
multi.lm <-
```

(@) Interpret one of the significant confidence intervals.

<br>

#### C. Prediction Intervals for a New Observational Unit
A prediction interval is a range of values that predicts the value of a new observation based on the existing model. We can calculate a prediction interval using the formula:
$$\small \bar{y_i }\pm \text{(Multiplier) (SE of Residuals)} \sqrt{1 + \tfrac{1}{n_i}}$$

<br>

(@) Estimate a prediction interval for average kilograms lost for an individual who is on the Keto diet. Interpret the prediction interval.
```{r, Prediction interval}
 
```

(@) When we are making a prediction where is our 'uncertainty' coming from?
