---
title: "Lesson 25 NFL Field Goals (continued) (6.3)"
author: "LTC J.K. Starling"
output:
  html_document:
    df_print: paged
---

```{r setup, message=FALSE, echo=TRUE, include=FALSE, warning=FALSE}
knitr::opts_chunk$set(
	echo = TRUE,
	message = FALSE,
	warning = FALSE
)
colorize <- function(x, color) {
  if (knitr::is_latex_output()) {
    sprintf("\\textcolor{%s}{%s}", color, x)
  } else if (knitr::is_html_output()) {
    sprintf("<span style='color: %s;'>%s</span>", color, 
      x)
  } else x
}
# `r colorize("....", "blue")`  # use this to color text
```

```{r}
# Load packages
library(tidyverse)

fg <- read.csv("https://raw.githubusercontent.com/jkstarling/MA376/main/fg_2021-2.csv",header=T, stringsAsFactors = TRUE)
fg <- fg %>%  mutate(field_goal_result = fct_rev(as.factor(field_goal_result)))

# setwd("C:/Users/james.starling/OneDrive - West Point/Teaching/MA376/JimsLessons_AY23-1/Block III/LSN25 (6.3)/")
# field_goal <- read.csv("fg_2021-2.csv", header = TRUE, stringsAsFactors = FALSE)


```

#### Background

(0) Last class we looked at the 2021 NFL Season and field goal kicks. In this lesson we will build our previous models.

We created a logistic regression model on the surface and distance.

```{r echo=TRUE, message=FALSE, warning=FALSE}
contrasts(fg$surface)
surface.glm <- fg %>% glm(field_goal_result ~ surface, data = ., family = 'binomial')
distance.glm <- glm(field_goal_result ~  kick_distance, data = fg, family = 'binomial')
summary(surface.glm)
summary(distance.glm)
```

\vspace{0.5in}

(1) We then took a look at the coefficients for each model and found the confidence intervals on those coefficients. How did we interpret these coefficients? What conclusions can we draw based on the confidence intervals?

```{r echo=TRUE, fig.height=3, fig.width=3, message=FALSE, warning=FALSE, fig.align='center'}
coef(surface.glm) 
confint(surface.glm)

coef(distance.glm)                
confint(distance.glm) 
```

**For the surface model, the intercept of 1.68 is the predicted log odds for a field goal on grass with a change in the log odds of -0.144 in the log odds if kicking on turf.**

**For the distance model, the intercept of 6.55 is the log odds of making a kick at 0 yards, with a change in the log odds of -0.11 for every increase of 1 yard distance.**

**The CI for the surface model DOES contain 0, indicating that there is not a statistically significant relationship between surface type and log odds. The CI for the distance model did NOT contain 0, indicating that there is a significant relationship between distance and log odds.**

\vspace{0.5in}


(2) We can also take the exponential of the coefficients and their CIs. Why would we do this? What is the interpretation of these results?

```{r}
exp(coef(surface.glm))           
exp(confint(surface.glm))

exp(coef(distance.glm))           
exp(confint(distance.glm))   
```


**We exponentiated so that we can get the odds-ratios for each of the explanatory variables.**

**We reach the same conclusion (regarding statistical significance), but we we are dealing with the odds changes based on surface type/distance. We should focus on whether or not the CI contains 1 now (instead of 0 with the log-odds).**

**The odds ratio for the surface now contains 1, so we cannot be certain that kicking from turf increase our odds (i.e. it could decrease by a factor or 0.866 or it could increase by a factor of 1.742). On the contrary, for the CI for the odds ratio of the distance slope coefficient tells us that we can (with some level of certainty) expect the odds to decrease by a factor of 0.874 to 0.912).**

\vspace{0.5in}


(3) Create a new logistic regression model that predicts the field goal result, taking into account the surface type and the kick distance. Discuss the results of this model.

```{r}
surf.dist.glm <- glm(field_goal_result ~ surface + kick_distance, data=fg, family='binomial')
summary(surf.dist.glm)
coef(surf.dist.glm)                
confint(surf.dist.glm)
```

**We see that we have a lower deviance (as expected) when compared to just the model with kick distance included, but the kick distance is statistically significant while the surface type is still not statistically significant. The intercept of 6.49 is the log odds of a kick on grass at 0 yards, while we see an increase of the log odds on turf of 0.144 and change of -0.113 in the log odds for each 1 yard increase in distance.**

\vspace{0.5in}


(4) Create a new logistic regression model that predicts the field goal result, taking into account the *interaction* between surface type and kick distance. Discuss the results of this model.

```{r}
int.glm <- glm(field_goal_result ~ surface * kick_distance, data=fg, family='binomial')
summary(int.glm)
coef(int.glm)                # Predicted log odds of making the field goal
confint(int.glm)
```


**We can see that only the kick_distance variable is statistically significant. The deviance has continued to decrease (but not much) while the AIC has risen. The interaction term shows that there is a slight increase in the log-odds as the distance increases while kicking off of turf. **

\vspace{0.5in}




(5) Calculate:

-   the overall percentage rate for field goals made
-   correct classification rate with `surface.lm`
-   correct classification rate with `distance.lm`
-   correct classification rate with `surf.dist.lm`
-   correct classification rate with `int.lm`

(What do we need to specify in order to calculate the percentages/rates? )

```{r}
calc.accuracy <- function(A){
  denom <- sum(A)
  TP <- 0; TN <- 0
  TP <- tryCatch(A["made","made.p"], silent=TRUE, error = function(ee) 0)
  TN <- tryCatch(A["missed","missed.p"], silent=TRUE, error = function(ee) 0)
  return ((TP + TN) / denom)
}

# calculate overall percentage
fg.attempts <- length(fg$field_goal_result) # should be 1076
fg.made <- sum(fg$field_goal_result == 'made')
overall.fg.rate <- fg.made / fg.attempts; 
overall.fg.rate

my.rate <- 0.5
# using surface model
fg.surf.pred <- ifelse(surface.glm$fitted.values >= my.rate, "made.p", "missed.p")
surf.tab <- table(fg$field_goal_result, fg.surf.pred)
surf.tab
fg.surf.rate <- calc.accuracy(surf.tab)
fg.surf.rate

# using distance model
fg.dist.pred <- ifelse(distance.glm$fitted.values >= my.rate, "made.p", "missed.p")
dist.tab <- table(fg$field_goal_result, fg.dist.pred)
dist.tab
fg.dist.rate <- calc.accuracy(dist.tab)
fg.dist.rate

# using surf + distance model
fg.sd.pred <- ifelse(surf.dist.glm$fitted.values >= my.rate, "made.p", "missed.p")
sd.tab <- table(fg$field_goal_result, fg.sd.pred)
fg.sd.rate <- calc.accuracy(sd.tab)
fg.sd.rate

# using surf * distance model
int.pred <- ifelse(int.glm$fitted.values >= my.rate, "made.p", "missed.p")
sd.tab <- table(fg$field_goal_result, int.pred)
int.rate <- calc.accuracy(sd.tab)
int.rate

c(fg.surf.rate, fg.dist.rate, fg.sd.rate, int.rate)
# cor(fg$field_goal_result, fg.sd.pred)

```


\vspace{0.5in}

(6) How can we check overall model performance?\


```{r}
my.rate <- 0.5
fgmade <- ifelse(fg$field_goal_result == 'made', 1, 0)
surf.made.p <- ifelse(unname(surface.glm$fitted.values) >= my.rate, 1, 0)
dist.made.p <- ifelse(unname(distance.glm$fitted.values) >= my.rate, 1, 0)
surfdist.made.p <- ifelse(unname(surf.dist.glm$fitted.values) >= my.rate, 1, 0)

c(sum(surf.made.p), sum(dist.made.p), sum(sum(dist.made.p)))
cor(fgmade, surf.made.p)^2
cor(fgmade, dist.made.p)^2
cor(fgmade, surfdist.made.p)^2

```

**We can use** $R^2$. The book mentions this on p. 481 that it is equal to the correlation between $cor(\hat y, y)$ but we have to convert the fg's made/missed and predicted fg's made/missed to 1's and 0's.

**You will have to choose the `my.rate` parameter carefully to avoid imbalances when calculating the correlation.**

    \vspace{0.5in}
    
    
(7) In R, redo  (5) and (6) above with various values for `my.rate`. What is the best choice for `my.rate` for maximizing the accuracy? for maximizing the $R^2$? 


```{r}
for (mythreshold in seq(0.5, 0.65, 0.01)){
# for (mythreshold in seq(0.05, 0.95, 0.05)){
  my.rate <- mythreshold

  # using distance model
  fg.dist.pred <- ifelse(distance.glm$fitted.values >= my.rate, "made.p", "missed.p")
  dist.tab <- table(fg$field_goal_result, fg.dist.pred)
  # dist.tab
  fg.dist.rate <- calc.accuracy(dist.tab)
  print(c(my.rate, fg.dist.rate))
}
```

